---
title: "Detecting Drug Synergy and Antagnoism with PharmacoGx 3.0"
author:
- name: Christopher Eeles
  affiliation: &pm Princess Margaret Cancer Centre, University Health Network, Toronto Canada
  email: christopher.eeles@uhnresearch.ca
- name: Feifei Li
  affiliation: &ccb Department of Cell & System Biology, University of Toronto, Toronto Canada
- name: Petr Smirnov
  affiliation:
    - *pm
    - &mbp Department of Medical Biophysics, University of Toronto, Toronto Canada
  email: psmirnov@utoronto.ca
- name: Benjamin Haibe-Kains
  affiliation:
    - *pm
    - *mbp
  email: benjamin.haibe.kains@utoronto.ca
output:
  BiocStyle::html_document
abstract: |
  Drug combinations are an effective strategy to over come incomplete response
  or acquired resistance to monotherapies in clinical oncology and beyond. As
  such, detecting molecular signatures of drug synergy or antagonism in
  pre-clinical model systems is a major priority to accelerate the adoption of
  novel treatment strategies in the clinic. The release of PharmacoGx 3.0
  introduces support for storing, analyzying and visualizing the results of drug
  combination experiments using the PharmacoSet class, and will enable
  reseachers across a range of disciplines to more easily mine the published
  drug combination literature for promising signatures of synergy or antagonism
  which can be validated in retrospective patient data and prioritized
  for prospective clinical trails.
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Getting started




# Style macros

_BiocStyle_ introduces the following macros for referring to _R_ packages:

* `r Biocpkg("IRanges")`, for _Bioconductor_ software, annotation and experiment data packages,
* `r CRANpkg("data.table")`, for _R_ packages available on CRAN,
* `r Githubpkg("rstudio/rmarkdown")`, for _R_ packages available on GitHub,
* `r Rpackage("MyPkg")`, for _R_ packages that are _not_ available on _Bioconductor_, CRAN or GitHub.


# Figures

Assign captions to figures in the code chunk option `fig.cap` to automatically number them, and to be able to reference them, see Figure \@ref(fig:plot). The figure label is generated from the code chunk label by prefixing it with `fig:`.

```{r plot, fig.cap="Regular figure. The first sentence of the figure caption is automatically emphasized to serve as figure title.", echo=FALSE}
plot(cars)
```

Small and wide figures can be specified by `fig.small` and `fig.wide` code chunk options.

```{r small, fig.cap="Small figure. A plot produced by a code chunk with option `fig.small = TRUE`.", fig.small=TRUE, echo=FALSE}
plot(cars)
```

```{r wide, fig.cap="Wide figure. A plot produced by a code chunk with option `fig.wide = TRUE`.", fig.wide=TRUE, echo=FALSE}
plot(cars)
```


# Equations

To number and reference equations, put them in equation environments and assign labels to them, see Equation \@ref(eq:binom).

\begin{equation}
  f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
  (\#eq:binom)
\end{equation}


# Tables

Like figures, tables with captions will also be numbered and can be referenced, see Table \@ref(tab:table).

Fruit   | Price
------- | -----
bananas | 1.2
apples  | 1.0
oranges | 2.5

: (\#tab:table) A simple table. With caption.


# Cross-references

Apart from referencing figures (Section \@ref(figures)), tables (Section \@ref(tables)), and equations (Section \@ref(equations)), you can also use the same syntax to refer to sections by their default labels generated by pandoc.


# Side notes

Footnotes are displayed as side notes on the right margin^[this is a side note entered as a footnote], which has the advantage that they appear close to the place where they are defined.


# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
